---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---

# DATA WRANGLING

### FINANCIAL AID DATA

```{r,warning=FALSE}
financial_aid<-read.csv("financial_aid.csv",header = T)
summary(financial_aid)

```

Combine by summing the loan, scholarship, grant and work/study data from 2011 to 2017

```{r,warning=FALSE}

library(tidyverse)
attach(financial_aid)
financial_aid=financial_aid %>%
  mutate(Total_loan = select(.,c(X2012.Loan,X2013.Loan,X2014.Loan,
           X2015.Loan,X2016.Loan,X2017.Loan)) %>%
           rowSums(na.rm = TRUE))

financial_aid<-financial_aid%>%
  mutate(Total_grant=select(.,c(X2012.Grant,X2013.Grant,X2014.Grant,
          X2015.Grant,X2016.Grant,X2017.Grant))%>%
           rowSums(na.rm = TRUE))

financial_aid<-financial_aid%>%
  mutate(Total_scholarship=select(.,c(X2012.Scholarship,X2013.Scholarship,X2014.Scholarship,
            X2015.Scholarship,X2016.Scholarship,X2017.Scholarship))%>%
           rowSums(na.rm = TRUE))

financial_aid<-financial_aid%>%
  mutate(Total_WorkStudy=select(.,c(X2012.Work.Study,X2013.Work.Study,
    X2014.Work.Study,X2015.Work.Study,X2016.Work.Study,X2017.Work.Study))%>%
           rowSums(na.rm = TRUE))

# Variables 
colnames(financial_aid)

```


Data cleaning : financial aid data
```{r,warning=FALSE}
# drop the extra periodical financial data 
financial_aid<-financial_aid[,-c(9:32)]

# Load the train labels and the TestIDs
train_labels=read.csv("DropoutTrainLabels.csv",header = T)
testIDs=read.csv("TestIDs.csv",header = T)

```


JOIN the train labels and the financial aid data

```{r,warning=FALSE}
library(dplyr)
financial_aid_train=left_join(train_labels,financial_aid,by="StudentID")
dim(financial_aid_train)
# JOIN the test IDs and the financial aid data
financial_aid_test=left_join(testIDs,financial_aid,by="StudentID")
dim(financial_aid_test)

```


Check for duplicated StudentID (TRUE=No duplicates, FALSE=duplicates present)

```{r,warning=FALSE}

length(unique(financial_aid_train$StudentID)) == nrow(financial_aid_train)

```


### STATIC DATA
STATIC data files loaded and merged 

```{r,warning=FALSE}
#load all files and merge them
myETL=function(mypath){
  filenames = list.files(path=mypath, full.names=TRUE)
  file_load = function(x){read.csv(file=x,header=T)}
  datalist = lapply(filenames, file_load)
  data2 = do.call(rbind, lapply(datalist, as.data.frame))
  return(data2)
}
# call the function
mergedStatic<-myETL("D:/KAZI/UPWORK/KAGGLE COMPETITION/Student Retention Challenge Data/Student Static Data")

# Variable names in static dataset
colnames(mergedStatic)

```


Check for duplicated StudentID (TRUE=No duplicates, FALSE=duplicates present)
```{r,warning=FALSE}
length(unique(mergedStatic$StudentID)) == nrow(mergedStatic)

```


JOIN the train labels and the static data

```{r,warning=FALSE}
library(dplyr)
static_train=left_join(train_labels,mergedStatic,by="StudentID")
summary(static_train)

# JOIN the test IDs and the static data
static_test=left_join(testIDs,mergedStatic,by="StudentID")

```



### PROGRESS DATA
Progress data files loaded and merged
```{r,warning=FALSE}

myETL=function(mypath){
filenames = list.files(path=mypath, full.names=TRUE)
file_load = function(x){read.csv(file=x,header=T)}
datalist = lapply(filenames, file_load)
data2 = do.call(rbind, lapply(datalist, as.data.frame))
return(data2)
}

# call the function
mergedProgress<-myETL("D:/KAZI/UPWORK/KAGGLE COMPETITION/Student Retention Challenge Data/Student Progress Data")

# variable names in the progress data
colnames(mergedProgress)
dim(mergedProgress)
```


Check for duplicated StudentID (TRUE=No duplicates, FALSE=duplicates present)
```{r,warning=FALSE}

length(unique(mergedProgress$StudentID)) == nrow(mergedProgress)


```


Summarise the data to get rid of duplicated StudentID

```{r,warning=FALSE}
library(plyr)
prog1=ddply(mergedProgress,.(StudentID),summarize,
            CompleteDevMath=mean(CompleteDevMath),CompleteDevEnglish=mean(CompleteDevEnglish),
            Major1=mean(Major1),Major2=mean(Major2),Complete1=mean(Complete1),
            Complete2=mean(Complete2),CompleteCIP1=mean(CompleteCIP1),
            CompleteCIP2=mean(CompleteCIP2),TransferIntent=mean(TransferIntent),
            DegreeTypeSought=mean(DegreeTypeSought),TermGPA=mean(CumGPA),
            CumGPA=mean(CumGPA),number=length(StudentID))
dim(prog1)

# drop the irrelevant frequency column
prog1<-prog1[,-14]
dim(prog1)


# JOIN the train labels and the progress data
library(dplyr)
progress_train=left_join(train_labels,prog1,by="StudentID")
# JOIN the test IDs and the progress data
progress_test=left_join(testIDs,prog1,by="StudentID")

dim(progress_train)

# variables names for the progress data
summary(progress_train)

```

Check to ensure that all the datasets are of the same sizes before merging them

```{r}

dim(financial_aid_train)
dim(financial_aid_test)

dim(static_train)
dim(static_test)

dim(progress_train)
dim(progress_test)

```



### TRAINING DATASET

```{r}

library(dplyr)
join1<-inner_join(financial_aid_train,static_train,by="StudentID")
TRAIN_DATA=inner_join(join1,progress_train,by="StudentID")
colnames(TRAIN_DATA)

dim(TRAIN_DATA)

# TESTING DATASET
join2<-inner_join(financial_aid_test,static_test,by="StudentID")
TEST_DATA<-inner_join(join2,progress_test,by="StudentID")
dim(TEST_DATA)


# Clean the train data by removing extra dropout variables (Dropout.y,Dropout)
TRAIN_DATA<-select(TRAIN_DATA,-c("Dropout.y"))
TRAIN_DATA<-select(TRAIN_DATA,-c("Dropout"))

# Rename the droput.x to just dropout for consistency
library(dplyr)
colnames(TRAIN_DATA)[colnames(TRAIN_DATA)=="Dropout.x"] <- "Dropout"

dim(TRAIN_DATA)
dim(TEST_DATA)

# The response variable
TRAIN_DATA$Dropout[1:5]

# The features of our project that will be analysed further
colnames(TEST_DATA)

```


```{r,warning=FALSE}
# convert empty spaces and (-1) to NULL in order to facilitate imputation later on
TRAIN_DATA[TRAIN_DATA=="-1"]<-NA
TRAIN_DATA[TRAIN_DATA==""]<-NA

# convert the theorically categorical variables to  factors 
names=c('DegreeTypeSought','TransferIntent','CompleteDevEnglish','CompleteDevMath',
  'GatewayEnglishStatus','GatewayMathStatus','EngPlacement','MathPlacement',
  'HighDeg','EnrollmentStatus','FirstGen','HSDip','BirthMonth','BirthYear',
  'Gender','Campus','CohortTerm','cohort_term','Dropout')

TRAIN_DATA[,names] <- lapply(TRAIN_DATA[,names] , factor)
str(TRAIN_DATA)

```



# EXPLORATORY DATA ANALYSIS

```{r}
# Summarize the data
summary(TRAIN_DATA)
# visualize the financial information in relation to the dropout variable
library(dplyr)
ggplot(data = TRAIN_DATA, mapping = aes(x =Dropout,y=Total_loan))+
  geom_boxplot()

ggplot(data = TRAIN_DATA, mapping = aes(x =Dropout,y=Total_grant))+
  geom_boxplot()

ggplot(data = TRAIN_DATA, mapping = aes(x =Dropout,y=Total_scholarship))+
  geom_boxplot()

ggplot(data = TRAIN_DATA, mapping = aes(x =Dropout,y=Total_WorkStudy))+
  geom_boxplot()

```


Frequency analysis for categorical variables in relation to the Dropout variable
```{r}
# Dropout
table(TRAIN_DATA$Dropout,exclude = NULL)


# Marital.Status
table(TRAIN_DATA$Marital.Status,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Marital.Status,fill=Dropout))

# Father.s.Highest.Grade.Level
table(TRAIN_DATA$Father.s.Highest.Grade.Level,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Father.s.Highest.Grade.Level,fill=Dropout))

# Mother.s.Highest.Grade.Level
table(TRAIN_DATA$Mother.s.Highest.Grade.Level,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Mother.s.Highest.Grade.Level,fill=Dropout))

# Housing
table(TRAIN_DATA$Housing,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Housing,fill=Dropout))


# Cohort
table(TRAIN_DATA$Cohort,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Cohort,fill=Dropout))

# CohortTerm
table(TRAIN_DATA$CohortTerm,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = CohortTerm,fill=Dropout))

# MathPlacement
table(TRAIN_DATA$MathPlacement,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = MathPlacement,fill=Dropout))


# State
table(TRAIN_DATA$State,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = State,fill=Dropout))

# Gender
table(TRAIN_DATA$Gender,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Gender,fill=Dropout))

# Hispanic
table(TRAIN_DATA$Hispanic,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Hispanic,fill=Dropout))

# AmericanIndian
table(TRAIN_DATA$AmericanIndian,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = AmericanIndian,fill=Dropout))


# Asian
table(TRAIN_DATA$Asian,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Asian,fill=Dropout))

# Black
table(TRAIN_DATA$Black,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = Black,fill=Dropout))

# NativeHawaiian
table(TRAIN_DATA$NativeHawaiian,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = NativeHawaiian,fill=Dropout))

# White
table(TRAIN_DATA$White,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = White,fill=Dropout))


# TwoOrMoreRace
table(TRAIN_DATA$TwoOrMoreRace,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = TwoOrMoreRace,fill=Dropout))


# HSDip
table(TRAIN_DATA$HSDip,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = HSDip,fill=Dropout))


# TwoOrMoreRace
table(TRAIN_DATA$TwoOrMoreRace,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = TwoOrMoreRace,fill=Dropout))

# HighDeg
table(TRAIN_DATA$HighDeg,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = HighDeg,fill=Dropout))

# EngPlacement
table(TRAIN_DATA$EngPlacement,TRAIN_DATA$Dropout,exclude = NULL)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = EngPlacement,fill=Dropout))

# GatewayMathStatus
table(TRAIN_DATA$GatewayMathStatus,TRAIN_DATA$Dropout)
ggplot(data = TRAIN_DATA) +
  geom_bar(mapping = aes(x = GatewayMathStatus,fill=Dropout))

```

Correlation matrix of the financial aid data to check for multicollinearity among the continuous independent variables

```{r}
# Load the GGally package
library(GGally)
# Create a scatter plot matrix
vars <- c("Total_loan", "Total_grant", "Total_WorkStudy", "Total_scholarship",
          "Adjusted.Gross.Income","Parent.Adjusted.Gross.Income","CumLoanAtEntry")
ggpairs(TRAIN_DATA[vars])

```


Correlation matrix of the student performance to check for multicollinearity among the continuous independent variables

```{r}

# Load the GGally package
library(GGally)
# Create a scatter plot matrix
vars <- c("CumGPA", "HSGPAWtd","TermGPA","HSGPAUnwtd","HSGPAWtd")
ggpairs(TRAIN_DATA[vars])

```



#### Hypotheses tests 

```{r}
# Test for association
# The dropout of students is not related to gender of a student
attach(TRAIN_DATA)
chisq.test(Dropout,Gender)

# The dropout of students is not related to Marital.Status of a student 
chisq.test(Marital.Status,Dropout)

# The dropout of students is not related to the housing of a student 
chisq.test(Housing,Dropout)

 #The dropout of students is not related to the Father.s.Highest.Grade.Level 
chisq.test(Father.s.Highest.Grade.Level,Dropout)

# The dropout of students is not related to the Mother.s.Highest.Grade.Level 
chisq.test(Mother.s.Highest.Grade.Level,Dropout)

# The dropout of students is not related to GatewayMathStatus of the student
chisq.test(GatewayEnglishStatus,Dropout)


```


### IMPUTATION OF MISSING VALUES

Summarize and look for the count of NA's in Individual variables 
```{r}
summary(TRAIN_DATA)
```

List the columns with missing values

```{r}
colnames(TRAIN_DATA)[colSums(is.na(TRAIN_DATA)) > 0]

```


Looking at the dataframe there are variables that have missing values close to 90% of the whole column (>11,000), these variables if imputed may introduce bias, hence they are dropped

```{r}
drop.cols <- c('HSGPAWtd', 'FirstGen','TransferIntent','Campus',
               'Address2','Major2')
TRAIN_DATA<-TRAIN_DATA %>% select(-drop.cols)
dim(TRAIN_DATA)
str(TRAIN_DATA)
```


Impute categorical variables with mode

```{r}

val<-unique(TRAIN_DATA$Marital.Status[!is.na(TRAIN_DATA$Marital.Status)]) # Values in vec_miss
mode <- val[which.max(tabulate(match(TRAIN_DATA$Marital.Status, val)))] # Mode of vec_miss
TRAIN_DATA$Marital.Status[is.na(TRAIN_DATA$Marital.Status)]<-mode # Impute by mode

val<-unique(TRAIN_DATA$Father.s.Highest.Grade.Level[!is.na(TRAIN_DATA$Father.s.Highest.Grade.Level)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Father.s.Highest.Grade.Level, val)))] 
TRAIN_DATA$Father.s.Highest.Grade.Level[is.na(TRAIN_DATA$Father.s.Highest.Grade.Level)]<-mode 

val<-unique(TRAIN_DATA$Mother.s.Highest.Grade.Level[!is.na(TRAIN_DATA$Mother.s.Highest.Grade.Level)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Mother.s.Highest.Grade.Level, val)))] 
TRAIN_DATA$Mother.s.Highest.Grade.Level[is.na(TRAIN_DATA$Mother.s.Highest.Grade.Level)]<-mode 

val<-unique(TRAIN_DATA$Housing[!is.na(TRAIN_DATA$Housing)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Housing, val)))] 
TRAIN_DATA$Housing[is.na(TRAIN_DATA$Housing)]<-mode


val<-unique(TRAIN_DATA$EngPlacement[!is.na(TRAIN_DATA$EngPlacement)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$EngPlacement, val)))] 
TRAIN_DATA$EngPlacement[is.na(TRAIN_DATA$EngPlacement)]<-mode

val<-unique(TRAIN_DATA$MathPlacement[!is.na(TRAIN_DATA$MathPlacement)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$MathPlacement, val)))] 
TRAIN_DATA$MathPlacement[is.na(TRAIN_DATA$MathPlacement)]<-mode

val<-unique(TRAIN_DATA$CompleteDevEnglish[!is.na(TRAIN_DATA$MathPlacement)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$CompleteDevEnglish, val)))] 
TRAIN_DATA$CompleteDevEnglish[is.na(TRAIN_DATA$CompleteDevEnglish)]<-mode

val<-unique(TRAIN_DATA$CompleteDevMath[!is.na(TRAIN_DATA$CompleteDevMath)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$CompleteDevMath, val)))] 
TRAIN_DATA$CompleteDevMath[is.na(TRAIN_DATA$CompleteDevMath)]<-mode

val<-unique(TRAIN_DATA$Address1[!is.na(TRAIN_DATA$Address1)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Address1, val)))] 
TRAIN_DATA$Address1[is.na(TRAIN_DATA$Address1)]<-mode 

val<-unique(TRAIN_DATA$State[!is.na(TRAIN_DATA$State)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$State, val)))] 
TRAIN_DATA$State[is.na(TRAIN_DATA$State)]<-mode 

val<-unique(TRAIN_DATA$City[!is.na(TRAIN_DATA$City)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$City, val)))] 
TRAIN_DATA$City[is.na(TRAIN_DATA$City)]<-mode 

val<-unique(TRAIN_DATA$Zip[!is.na(TRAIN_DATA$Zip)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Zip, val)))] 
TRAIN_DATA$Zip[is.na(TRAIN_DATA$Zip)]<-mode 

val<-unique(TRAIN_DATA$BirthYear[!is.na(TRAIN_DATA$BirthYear)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$BirthYear, val)))] 
TRAIN_DATA$BirthYear[is.na(TRAIN_DATA$BirthYear)]<-mode 


val<-unique(TRAIN_DATA$Hispanic[!is.na(TRAIN_DATA$Hispanic)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Hispanic, val)))] 
TRAIN_DATA$Hispanic[is.na(TRAIN_DATA$Hispanic)]<-mode 

val<-unique(TRAIN_DATA$AmericanIndian[!is.na(TRAIN_DATA$AmericanIndian)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$AmericanIndian, val)))] 
TRAIN_DATA$AmericanIndian[is.na(TRAIN_DATA$AmericanIndian)]<-mode 

val<-unique(TRAIN_DATA$Asian[!is.na(TRAIN_DATA$Asian)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Asian, val)))] 
TRAIN_DATA$Asian[is.na(TRAIN_DATA$Asian)]<-mode 

val<-unique(TRAIN_DATA$Black[!is.na(TRAIN_DATA$Black)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$Black, val)))] 
TRAIN_DATA$Black[is.na(TRAIN_DATA$Black)]<-mode 

val<-unique(TRAIN_DATA$NativeHawaiian[!is.na(TRAIN_DATA$NativeHawaiian)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$NativeHawaiian, val)))] 
TRAIN_DATA$NativeHawaiian[is.na(TRAIN_DATA$NativeHawaiian)]<-mode 

val<-unique(TRAIN_DATA$White[!is.na(TRAIN_DATA$White)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$White, val)))] 
TRAIN_DATA$White[is.na(TRAIN_DATA$White)]<-mode 

val<-unique(TRAIN_DATA$TwoOrMoreRace[!is.na(TRAIN_DATA$TwoOrMoreRace)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$TwoOrMoreRace, val)))] 
TRAIN_DATA$TwoOrMoreRace[is.na(TRAIN_DATA$TwoOrMoreRace)]<-mode 

val<-unique(TRAIN_DATA$HSDip[!is.na(TRAIN_DATA$HSDip)])
mode<-val[which.max(tabulate(match(TRAIN_DATA$HSDip, val)))] 
TRAIN_DATA$HSDip[is.na(TRAIN_DATA$HSDip)]<-mode 


# which columns have missing data
colnames(TRAIN_DATA)[colSums(is.na(TRAIN_DATA)) > 0]

```



Imputing continuous variables

First check for normal distribution in order to impute with mean or median if they are not normally distributed (bell-shaped histogram and density plot shows normality)

```{r,warning=FALSE}

#Density plot
library(ggpubr)
ggdensity(TRAIN_DATA$Adjusted.Gross.Income, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")
```


```{r,warning=FALSE}
#Histogram

hist(TRAIN_DATA$Adjusted.Gross.Income) #histogram
```


```{r,warning=FALSE}
# QQplot
ggqqplot(TRAIN_DATA$Adjusted.Gross.Income) #qqplot


```


```{r,warning=FALSE}
# Density plot
ggdensity(TRAIN_DATA$Parent.Adjusted.Gross.Income, 
          main = "Density plot of tooth length",
          xlab = "Tooth length")
```


```{r,warning=FALSE}
# Histogram
hist(TRAIN_DATA$Parent.Adjusted.Gross.Income) #histogram

```


```{r,warning=FALSE}
# QQplot
ggqqplot(TRAIN_DATA$Parent.Adjusted.Gross.Income) #qqplot


```

- NOTE: the continous variables did not display normal distribution properties, thus imputation will be done using median

#### Impute continuous variables using median

```{r,warning=FALSE}

for(i in 1:nrow(TRAIN_DATA)) {
  TRAIN_DATA[i, ][is.na(TRAIN_DATA[i,])]<-median(as.numeric(TRAIN_DATA[i,]), na.rm = TRUE)
}

```



# FEATURE ENGINEERING

Implement feature selection using Boruta package 
```{r,warning=FALSE}
library(Boruta)

# Now is the time to implement and check the performance of boruta package. The syntax of boruta is almost similar to regression (lm) method.

set.seed(123)
boruta.train <- Boruta(Dropout~.-StudentID, data = TRAIN_DATA, doTrace = 2)
print(boruta.train)


# Now, we’ll plot the boruta variable importance chart.
plot(boruta.train, xlab = "", xaxt = "n")
 lz<-lapply(1:ncol(boruta.train$ImpHistory),function(i)
  boruta.train$ImpHistory[is.finite(boruta.train$ImpHistory[,i]),i])
 names(lz) <- colnames(boruta.train$ImpHistory)
 Labels <- sort(sapply(lz,median))
 axis(side = 1,las=2,labels = names(Labels),
       at = 1:ncol(boruta.train$ImpHistory), cex.axis = 0.7)


```




```{r,warning=FALSE}
# Blue boxplots correspond to minimal, average and maximum Z score of a shadow attribute. Red, yellow and green boxplots represent Z scores of rejected, tentative and confirmed attributes respectively.
# Now is the time to take decision on tentative attributes.  The tentative attributes will be classified as confirmed or rejected by comparing the median Z score of the attributes with the median Z score of the best shadow attribute. Let’s do it.

final.boruta <- TentativeRoughFix(boruta.train)
print(final.boruta)


```




```{r,warning=FALSE}
# We’ll create a data frame of the final result derived from Boruta.
boruta.df <- attStats(final.boruta)
class(boruta.df)

print(boruta.df)


# It’s time for results now. Let’s obtain the list of confirmed attributes
features=getSelectedAttributes(final.boruta, withTentative = F) 
features

TRAIN_SET=TRAIN_DATA[,features]
TRAIN_SET$Dropout=TRAIN_DATA$Dropout
dim(TRAIN_SET)

```

Let’s understand the parameters used in Boruta as follows:

- maxRuns: maximal number of random forest runs.
You can consider increasing this parameter if tentative attributes are left.Default is 100.
- doTrace: It refers to verbosity level. 0 means no tracing. 1 means reporting attribute decision as soon as it is cleared. 2 means all of 1 plus additionally reporting each iteration. Default is 0.
- holdHistory: The full history of importance runs is stored if set to TRUE
(Default). Gives a plot of Classifier run vs. Importance when the plotImpHistory function is called.









