---
title: "Student Dropout Competition: Modelling process"
author: "Hamed"
date: "3/25/2020"
output: word_document
---

# Modelling


```{r,warning=FALSE}
# Load the engineered and transformed features
transformed_train=read.csv("transformed_train.csv",header = T)
transformed_test=read.csv("transformed_test.csv",header = T)

# Convert the response variable from an integer to a factor
transformed_train$Dropout=as.factor(transformed_train$Dropout)

# Load required packages
library(class)
library(caret)
library(rpart)
library(dplyr)


# Split the transformed_train dataset into training and validation datasets
set.seed(123)
## 80% of the sample size
smp_size <- floor(0.80 * nrow(transformed_train))
train_ind <- sample(seq_len(nrow(transformed_train)), size = smp_size)
train.set <- transformed_train[train_ind, ]
validation.set <- transformed_train[-train_ind, ]

dim(train.set)

str(train.set)
```




##  Gradient Boosting Models

GBM model training

```{r,warning=FALSE}
# Stochastic Gradient Boosting GBM model

library(caret)
library(gbm)

set.seed(123)

fit.gbm <- train(Dropout~., data=train.set, method="gbm", metric="Accuracy", trControl=trainControl(method="repeatedcv", number=10, repeats=3), verbose=FALSE)

fit.gbm

```


GBM: Model Evaluation and Prediction

```{r}
# Evaluation of model accuracy
predict_gbm<-predict.train(object=fit.gbm,validation.set,type="raw")

confusionMatrix(predict_gbm,validation.set$Dropout)
```


```{r}
# Predict new data
pred.gbm<-predict.train(object=fit.gbm,transformed_test,type="raw")
pred.gbm=as.data.frame(pred.gbm)

head(pred.gbm)
# save the prediction 
#write.csv(pred.gbm,"D:/Hamed/KAGGLE COMPETITION/FEATURES/my_new_submission/submission_gbm.csv")

```




##  Logistic Regression

```{r,warning=FALSE}
library(mlbench)     # for PimaIndiansDiabetes2 dataset
library(dplyr)       # for data manipulation (dplyr) 
library(broom)       # for making model summary tidy
library(visreg)      # for potting logodds and probability 
library(margins)     # to calculate Average Marginal Effects
library(rcompanion)  # to calculate pseudo R2
library(ROCR)        # to compute and plot Reciever Opering Curve

#Fitting a binary logistic regression
model_logi <- glm(Dropout~., data = train.set, family = "binomial")
#Model summary
summary(model_logi)

```

Model fit statistics

```{r,warning=FALSE}

# Pseudo R_squared values and Likelyhood ratio test
nagelkerke(model_logi)

```


ODDS Ratios

```{r,warning=FALSE}
# The ODDS ratio can be retrieved in a beautiful tidy formatted table
# using the tidy( ) function of broom package.
tidy(model_logi, exponentiate = TRUE, conf.level = 0.95) #odds ratio
```

Model Evaluation on Test Data Set

```{r,warning=FALSE}

# Confusion matrix
# predict the test dataset
pred <- predict(model_logi, validation.set, type="response") 
predicted <-ifelse(pred>0.5,1,0) # round of the value; >0.5 will convert to 1 else 0
table(predicted)

# Creating a contigency table
tab <- table(Predicted = predicted, Reference = validation.set$Dropout)
tab

```


Accuracy

```{r,warning=FALSE}
# Creating a dataframe of observed and predicted data
library(yardstick)
act_pred <- data.frame(observed = validation.set$Dropout, predicted=factor(predicted))

# Calculating Accuracy
accuracy_est <- accuracy(act_pred, observed, predicted)
print(accuracy_est)

```


Classification Report

```{r,warning=FALSE}
# Precision, F1-score and recall values
library(yardstick)
# Creating a actual/observed vs predicted dataframe
act_pred <- data.frame(observed = validation.set$Dropout, predicted =  
                         factor(predicted))
# Calculating precision, recall and F1_score
prec <- precision(act_pred, observed, predicted)
rec <- recall(act_pred, observed, predicted)
F1_score <- f_meas(act_pred, observed, predicted) #called f_measure
print(prec)
print(rec)
print(F1_score)

```


Prediction using new data

```{r,warning=FALSE}
pred_log <- predict(model_logi, transformed_test, type="response") 

predicted <- round(pred_log) # round of the value; >0.5 will convert to 1 else 0
submission_lr=as.data.frame(predicted)

head(submission_lr)

# save the file
#write.csv(submission_lr,"D:/Hamed/KAGGLE COMPETITION/FEATURES/my_new_submission/submission_lr.csv")


```


##  Support Vector Machine (SVM)

```{r,warning=FALSE}
library(dplyr)
library(mlr)
library(caret)
library(ROCR)
library(kernlab)
library(e1071)
library(foreach)
library(doParallel)
model_svm <- svm(Dropout~.,data=train.set,trControl=trainControl("cv",number=10),
                 tuneGrid = expand.grid(C=c(.01,.02,.05,.1,.2,.5,1,2,5,10)
                  ,degree=c(1:5),scale=c(0.01:1)),tuneLength = 4)

summary(model_svm)

model_svm$cost#displays cost , error, degree and scale of the model
model_svm$epsilon #displays the accuracy of the model crossvalidated 


```

Model Evaluation

```{r,warning=FALSE}

x=validation.set[,-53]
y=validation.set[,53]
pred <- predict(model_svm,x)
confusionMatrix(pred,y)


```

Predict the test data

```{r,warning=FALSE}
pred_test=predict(model_svm,transformed_test)
table(pred_test)

# Create a submission file
submission_svm=as.data.frame(pred_test)

head(submission_svm)
#write.csv(pred_test,"D:/Hamed/KAGGLE COMPETITION/FEATURES/my_new_submission/submission_svm.csv")

```



##  Decision Tree Model

```{r,warning=FALSE}
# load libraries
library(rpart)
library(rattle)
library(mlr) 
library(FSelector) 
library(rpart.plot)
```

First we have to make a classification task with our training set. This is where we can define which type of machine learning problem we’re trying to solve and define the target variable
```{r,warning=FALSE}

(dt_task <- makeClassifTask(data=train.set, target="Dropout"))

```

After creating a classification task we need to make a learner that will later take our task to learn the data. I have chosen the rpart decision tree algorithm. This is the Recursive Partitioning Decision Tree.

```{r,warning=FALSE}
(dt_prob <- makeLearner('classif.rpart', predict.type="prob"))

```

Hyper Parameter Tuning

```{r,warning=FALSE}
getParamSet("classif.rpart")

dt_param <- makeParamSet( makeDiscreteParam("minsplit", values=seq(5,10,1)),
            makeDiscreteParam("minbucket",values=seq(round(5/3,0), round(10/3,0), 1)),
            makeNumericParam("cp",lower = 0.01, upper = 0.05),
            makeDiscreteParam("maxcompete",
            values=6), makeDiscreteParam("usesurrogate", values=0),
            makeDiscreteParam("maxdepth", values=10) )

```

Optimization Algorithm

```{r,warning=FALSE}
ctrl = makeTuneControlGrid()

# Evaluating Tuning with Resampling
rdesc = makeResampleDesc("CV", iters = 3L, stratify=TRUE)

```

We can now use tuneParams to show us what combination of hyperparameter values as specified by us will give us the optimal result.

```{r,warning=FALSE}
set.seed(1000) 
(dt_tuneparam <- tuneParams(learner=dt_prob,resampling=rdesc,measures=list(tpr,auc,
            fnr, mmce, tnr, setAggregation(tpr, test.sd)), 
            par.set=dt_param,control=ctrl,task=dt_task,show.info = TRUE) )

```

Optimal HyperParameters

```{r,warning=FALSE}
list( 'Optimal HyperParameters' = dt_tuneparam$x, 
      'Optimal Metrics' = dt_tuneparam$y )

dtree <- setHyperPars(dt_prob, par.vals = dt_tuneparam$x)

```

Model Training

```{r}
set.seed(1000) 
dtree_train <- train(learner=dtree, task=dt_task) 
getLearnerModel(dtree_train)

rpart.plot(dtree_train$learner.model, roundint=FALSE, varlen=3, type = 3,
           clip.right.labs = FALSE, yesno = 2)
rpart.rules(dtree_train$learner.model, roundint = FALSE)

```

Model Prediction (Testing):
We now pass the trained learner to be used to make predictions with our test data.

```{r}

set.seed(1000) 
(dtree_predict <- predict(dtree_train, newdata = validation.set))

# The threshold for classifying each row is 50/50. This is by default
# but can be changed later (which I will do).
dtree_predict %>% calculateROCMeasures()

```


```{r}
dtree_predict.test <- predict(dtree_train, newdata = transformed_test)

sub_tree=as.data.frame(dtree_predict.test)
head(sub_tree)

#write.csv(sub_tree,"D:/Hamed/KAGGLE COMPETITION/FEATURES/my_new_submission/sub_tree.csv")


```



##  K-Nearest Neighbors (KNN)

```{r}
library(class)
library(caret)
library(rpart)
library(dplyr)

```


Create the train and validation labels

```{r,warning=FALSE}
train.dropout_labels <- train.set$Dropout
val.dropout_labels <-validation.set$Dropout


# Lets get a good k value

i=1
k.optm=1
for (i in 1:28){
  knn.mod <- knn(train=train.set, test=validation.set, cl=train.dropout_labels, k=i)
  k.optm[i] <- 100 * sum(val.dropout_labels == knn.mod)/NROW(val.dropout_labels)
  k=i
  cat(k,'=',k.optm[i],'
')
}

# Plot the K to check Accuracy plot for best k values
plot(k.optm, type="b", xlab="K- Value",ylab="Accuracy level")

```

Use the best K value to fit a model to the data

```{r,warning=FALSE}
library(kknn)
knn.fit <- train.kknn(as.factor(Dropout)~., train.set, ks = 22,
                      kernel = "rectangular", scale = TRUE)
pred.train.kknn <- predict(knn.fit, validation.set)

# Lets look at the performance metrics
confusionMatrix(table(pred.train.kknn ,val.dropout_labels))

```



Prediction using a new data

```{r,warning=FALSE}

library(caret)

pred_knn<-predict(object=knn.fit,transformed_test,type="raw")
pred_knn=as.data.frame(pred_knn)


submission_knn=as.data.frame(pred_knn)
head(submission_knn)

# save the file
#write.csv(submission_knn,"D:/Hamed/KAGGLE COMPETITION/FEATURES/my_new_submission/submission_knn.csv")

```


