---
title: "Data Wrangling - 3.3"
author: "HILLARY"
date: "2/23/2020"
output: word_document
---

## (a).	Load the datasets/tables flights and airlines

```{r, warning=FALSE}

# (a).	Install and load the library nycflights13
library(tidyverse)
library(dplyr)
library(nycflights13)

```



# (b).	Load the datasets/tables flights and airlines

```{r, warning=FALSE}

airlines=data.frame(airlines)
flights=data.frame(flights)
airports=data.frame(airports)
planes=data.frame(planes)
weather=data.frame(weather)

head(airlines)
head(flights)
head(airports)
head(planes)
head(weather)

```


## (c). What is the primary key of the planes table 

```{r, warning=FALSE}

#Planes$tailnum this is the primary key as it uniquely identifies each plane in the planes table
planes%>%count(tailnum)%>%filter(n>1)


```


## (d).	Add full airline name from the airlines table to the flights table that keeps all the records in the flights table by using the appropriate join

```{r, warning=FALSE}

flights_with_names<-left_join(x=flights,y=airlines,by='carrier',all.x=TRUE)

head(flights_with_names)


```



## (e).	Now add the destination latitude and longitude to the flights table from the airports table by using the appropriate join

```{r, warning=FALSE}


# Rename "faa" to "dest" in order to use it as joining key
names(airports)[names(airports)=="faa"] <- "dest"

flights_dest<-left_join(x=flights,y=airports,by='dest',all.x=TRUE)[c("year","month",
              "day","dep_time","sched_dep_time","dep_delay","arr_time",
              "sched_arr_time","arr_delay","carrier","flight","tailnum","origin",
              "dest","air_time" ,"distance","hour","minute","time_hour","lat","lon")]

head(flights_dest)

```


