---
title: "Homework 3.2 - Resampling"
author: "HILLARY"
date: "2/23/2020"
output: word_document
---


```{r ,warning=FALSE}
library(caret)
library(rsample)   
library(purrr)
library(dplyr)
library(ggplot2)
library(scales)
library(mlbench)
library(kernlab)
library(sessioninfo)
theme_set(theme_bw())
library(RCurl)

```


## Get the German Credit dataset from the UCI machine learning repository

```{r ,warning=FALSE}

UCI_german<-getURL("http://archive.ics.uci.edu/ml/machine-learning-databases/statlog/german/german.data")

names<-c('checking_account','month','Credit_history','Purpose',
         'Credit_amount','Savings','employment','Installment_rate',
         'status','Other_debtors','Present_residence_since','Property',
         'Age','installment_plans','Housing','#credits','Job','#people',
         'Telephone','foreign','Cost_Matrix')

credit_data<-read.table(textConnection(UCI_german),sep=" ",
                        col.names=names)
str(credit_data)

```



# Fit a SVM model to predict the type of credit (good or bad) with the following resampling techniques:


## 5 fold cross validation

```{r ,warning=FALSE}

# convert class variable into factor 
credit_data$Cost_Matrix=as.factor(credit_data$Cost_Matrix)

# First split the dataset , 85% Training data ad 15% testing data
set.seed(123)
smp_size<- floor(0.85 * nrow(credit_data))
train_ind<-sample(seq_len(nrow(credit_data)), size = smp_size)
train.credit<-credit_data[train_ind, ]
test.credit<-credit_data[-train_ind, ]


dim(train.credit) #shape of the train data 

dim(test.credit) #shape of test data


#Modelling
train_control <- trainControl(method="cv", number=5)
model <- train(Cost_Matrix~., data=train.credit, trControl=train_control, method="svmLinear")
# summarize results
print(model)

# prediction
pred=predict(model,newdata = test.credit)
pred

# Show the accuracy for the training dataset
model$results

# Show the accuracy results for each resample
model$resample

# Show the accuracy for the test dataset
mean(pred == test.credit$Cost_Matrix)

```


## 10 fold cross-validation with 3 repeats


```{r ,warning=FALSE}

# define training control
train_control2 <- trainControl(method="repeatedcv", number=10, repeats=3)
model2<-train(Cost_Matrix~.,data=train.credit,trControl=train_control2, method="svmLinear")

# summarize results
print(model2)
pred2=predict(model2,newdata = test.credit)
pred2


# Show the accuracy for the training dataset
model2$results

# Show the accuracy results for each resample
model2$resample

# Show the accuracy for the test dataset
mean(pred2 == test.credit$Cost_Matrix)

```

## Leave group out with 5 iterations and 85% data used for training define training control

```{r setup,warning=FALSE}
# Modelling
train_control3 <- trainControl(method="LOOCV",number = 5)
model3<-train(Cost_Matrix~.,data=train.credit,trControl=train_control3, method="svmLinear")

# summarize results
print(model3)


# Prediction
pred3=predict(model3,newdata=test.credit)
pred3


# Show the accuracy for the training dataset
model3$results

# Show the accuracy results for each resample
model3$resample

# Show the accuracy for the test dataset
mean(pred3 == test.credit$Cost_Matrix)


```



## Bootstrap with 25 iterations

```{r,warning=FALSE}

# Modeling
train_control4 <- trainControl(method="boot", number=25)
model4 <- train(Cost_Matrix~.,data=train.credit,trControl=train_control4,method="svmLinear")

# summarize results
print(model4)

# Prediction
pred4=predict(model4,newdata=test.credit)
pred4

# Show the accuracy for the training dataset
model4$results

# Show the accuracy results for each resample
model4$resample

# Show the accuracy for the test dataset
mean(pred4 == test.credit$Cost_Matrix)


```

# Which resampling method gives the best estimate of the test error?

- Selected the method with the highest test accuracy as it was the best 

```{r}
#5 Fold cross-validation
mean(pred == test.credit$Cost_Matrix)

# 10 fold cross-validation
mean(pred2 == test.credit$Cost_Matrix)


# Leave group out with 5 iterations and 85% data used for training define training control
mean(pred4 == test.credit$Cost_Matrix)



# Bootstrap with 25 iterations
mean(pred4 == test.credit$Cost_Matrix)


```

